Expression QTL hotspot dissection

### 2019-08-03 analysis/paper/Rnw/paper-thesis.Rnw
organized analysis/paper dir

\documentclass[oneside]{book}\usepackage[]{graphicx}\usepackage[]{color}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[utf8]{inputenc}
%\usepackage{lineno}
%\linenumbers

\usepackage{bbm}
\usepackage{setspace}
\usepackage[left = 1in, right = 1in, top = 1.5in, bottom = 1in]{geometry}
\usepackage{multirow}
\usepackage[framemethod=TikZ]{mdframed}
\usepackage{float}
\usepackage{newfloat}
\usepackage{multicol}
%\usepackage[lite]{mtpro2} %contains \widetilde
\usepackage[titletoc]{appendix}
\usepackage{amsmath}
\usepackage{hyperref}
\hypersetup{unicode=true,
%            pdftitle={Chapter 1},
%            pdfauthor={Frederick Boehm},
%            pdfborder={0 0 0},
            colorlinks=true,
            breaklinks=true}

\usepackage[
    backend=biber,
    style=chicago-authordate,
%    citestyle=authoryear,
    natbib=true, %citet and citep
    url=true,
    doi=true,
    eprint=true,
    maxbibnames=10,
    backref=true
]{biblatex}
\DeclareLanguageMapping{english}{english-apa}
\addbibresource{research.bib}
\usepackage{graphicx}
\usepackage{color, colortbl}
\definecolor{LightCyan}{rgb}{0.88,1,1}
\usepackage{subcaption}
\usepackage{array}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{longtable}
\usepackage[outdir=./figs/]{epstopdf}
\usepackage{etoolbox} % for messing with page style & title pages
\usepackage{fancyhdr}
\setlength{\headheight}{15.2pt}
\pagestyle{fancyplain}
\fancyhf{}
\lhead{}
\chead{}
%\rhead{}
\rhead{ \fancyplain{}{\thepage} } % put page number in upper right corner
\renewcommand{\headrulewidth}{0pt} % remove top line
\renewcommand{\footrulewidth}{0pt} % remove bottom line
\renewcommand{\headrule}{}
\title{The title}
\author{An Author}
\patchcmd{\titlepage}{empty}{fancy}{}{}
\patchcmd{\chapter}{plain}{fancy}{}{}
\DeclareCaptionType{equ}[][]
%https://stackoverflow.com/questions/149479/adding-a-caption-to-an-equation-in-latex
%\captionsetup[equ]{labelformat=empty}

\DeclareFloatingEnvironment[fileext=frm,placement={!ht},name=Frame]{myfloat}
\captionsetup[myfloat]{labelfont=bf}
\newenvironment{frameenv}[1]
    {\begin{myfloat}[tb]
    \begin{mdframed}[roundcorner=10pt,backgroundcolor=blue!10]
    \caption{#1}
    %\begin{multicols*}{2}
    }
    {%\end{multicols*}
    \end{mdframed}\end{myfloat}
    }

\title{}
\author{Frederick J. Boehm, Brian S. Yandell, and Karl W. Broman}
\date{\today}

%% from vignette tex file
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}

%% end from vignette tex file
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\doublespacing
\begin{titlepage}\thispagestyle{empty}
\tableofcontents
%\listoffigures
%\listoftables
%\pagestyle{headings} % page numbers appear at bottom of page, centered!
%\markright{\hfill}
\chapter{Introduction}

Genetics studies in model organisms like mice can identify
genomic regions that affect complex traits, such as systolic blood pressure and body weight \citep{sax1923association,soller1976power,lander1989mapping,broman2009guide,jansen2007quantitative}. These genomic regions are called ``quantitative trait loci'' or ``QTL''.
A genome-wide QTL ``scan'' reveals
associations between genotypes and phenotypes by considering each position,
one at a time, as a candidate QTL for the trait of interest.
A region with strong evidence of association with a complex trait, then, defines a QTL (for that trait).
Because nearby markers have correlated
genotypes, a QTL in a two-parent cross often spans multiple megabases in length
and may contain more than a hundred genes.
Identification of the causal gene (for a given complex trait) from among those genes near the QTL is
challenging and may require costly and time-consuming experiments.
The growing need for greater QTL mapping resolution fueled development over the
last two decades of model organism multiparental populations for high-resolution QTL mapping \citep{de2017back,churchill2004collaborative,svenson2012high,huang2012multiparent,shivakumar2018soybean,huang2011analysis,kover2009multiparent,tisne2017identification,stanley2017genetic}.
With experimentalists now measuring tens of thousands of biomolecular traits
in multiparental populations, the systems genetics community needs multivariate
statistical tools to fully examine the large volumes of data \citep{keller2018genetic,chick2016defining}.
Identifying loci that affect multiple traits can aid in identifying biomolecular interactions and in clarifying complex trait genetic architecture. A QTL that affects more than one trait is called a ``pleiotropic'' QTL.
A test of pleiotropy vs. separate QTL is one multivariate statistical tool that will inform
complex trait genetics by enabling researchers to identify the number of
unique QTL in a genomic region of interest.
In this thesis, I develop a test of pleiotropy vs. separate QTL in multiparental
populations. I study its statistical properties and demonstrate its utility
by analyzing experimental data.

Due to the breeding scheme, each Diversity Outbred mouse is a highly
heterozygous and essentially unique mosaic of
founders' DNA. Figure~\ref{fig:do} presents color-coded chromosomes for a collection of
subjects from a multiparental population like the Diversity Outbred mice. DNA from each founder
line is color-coded. Each pair of vertical bars corresponds to one subject. Each row corresponds to a single generation, with generation labels on the left. Generation number increases from top to bottom.
With each new generation, the Diversity Outbred mouse genomes are shuffled
in smaller and smaller chunks.

@@

One then fits a linear mixed effects model and calculates the log$_{10}$ likelihood at each position. As in two-parent crosses (Section~\ref{sec:qtl-two-parent}), one summarizes evidence for a QTL by plotting LOD scores across the genome. Because multiple hypothesis tests are performed, one typically wants to control family-wise error rate. One strategy to do this involves a permutation test, much like that in two-parent crosses \citep{churchill1994empirical}.

@@

We continue by stating what it means for one trait to mediate a relationship between a
DNA variant and another trait. To clarify our discussion, we refer to an example from
\citet{chick2016defining} (Figure~\ref{fig:Dhtkd1}). \citet{chick2016defining}, in
studying livers of 192 Diversity Outbred mice, found evidence that \emph{Dhtkd1}
transcript levels associated with a Chromosome 2 marker near the \emph{Dhtkd1} gene.
They also found that
the same marker affected DHTKD1 protein concentrations. (Note that DHTKD1 protein is the
product of translation of \emph{Dhtkd1} transcripts.)
As anticipated, mediation analysis, in which the DHTKD1 protein concentrations are
regressed on founder allele dosages (at the Chromosome 2 marker) demonstrated that
\emph{Dhtkd1} transcript levels act as a mediator between DHTKD1 protein concentrations
and founder allele dosages. In fact, the extent of the reduction in association
strength indicates that the primary pathway by which the genetic marker affects DHTKD1
protein concentrations is through \emph{Dhtkd1} transcript levels.

\begin{figure}
  \centering
  \includegraphics[width = 0.7\textwidth]{../QTLfigs/Figs/central_dogma.pdf}
  \caption{Biological information is encoded in DNA. This information is passed, via transcription, to sequence-specific RNA molecules. The process of translation transmits the information to sequence-specific proteins.}\label{fig:dogma}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width = 0.7\textwidth]{../QTLfigs/Figs/central_dogma-Dhtkd1.pdf}
  \caption{A DNA variant in the \emph{Dhtkd1} gene affects \emph{Dhtkd1} transcript abundances which, in turn, affect DHTKD1 protein concentrations.}\label{fig:Dhtkd1}
\end{figure}

@@

\begin{frameenv}{Four assumptions for causal inference}\label{frame1}
  \begin{enumerate}
\item No unmeasured confounding of the DNA variant-nonlocal transcript levels relationship
\item No unmeasured confounding of the local transcript levels-nonlocal transcript levels relationship
\item No unmeasured confounding of the DNA variant-local transcript levels relationship
\item No local transcript levels-nonlocal transcript levels confounder that is affected by the DNA variant
\end{enumerate}

\end{frameenv}

Our first statistical challenge is due to confounding. To claim that the LOD
difference statistic reflects a causal relationship, four assumptions about
confounding are needed (Frame~\ref{frame1}) \citep{vanderweele2015explanation},
yet
it is often difficult or impossible to recognize unmeasured confounders. Unmeasured
confounders, as I use the term here, are unmeasured random variables that confound a causal
relationship between two other random variables.
For example, Figure~\ref{fig:confounding} shows a directed graph. Arrows indicate causal
directions. For example, ``QTL'' causes ``M''. ``M'' designates the mediator (local transcript). ``Y'' denotes the nonlocal transcript
level in this setting, while ``C1'', ``C2'', and ``C3'' are covariates that confound,
respectively, the DNA variant - nonlocal transcript level relationship, the local transcript
level - nonlocal transcript level relationship, and the DNA variant - local transcript level
relationship. Failure to measure C1, C2, or C3 leads to violations of Assumption 1,
Assumption 2, or Assumption 3, respectively (Frame~\ref{frame1}).

\begin{figure}
\includegraphics{figs/confounding.pdf}
\caption{Example causal diagram showing relationships between random variables in an expression QTL study.}\label{fig:confounding}
\end{figure}

In studies of Diversity Outbred mice, relatedness is a possible confounder,
yet the linear models (Equations~\ref{model1},~\ref{model2},~\ref{model3},~\ref{model4}
in Frame~\ref{frame3}) fail to account for the complex relatedness patterns
among Diversity Outbred mice.
In fact, the only covariates in our models are for wave membership and sex.
The 400 mice were shipped from the Jackson Laboratory in four waves of 100 mice.
Because each wave was processed at a different time, I include wave membership
as a covariate with the goal of accounting for wave-related effects.
Other sources of confounding, such as batch effects in the phenotyping, may be
unmeasured.

@@

We designed a study to examine the possible roles for mediation analysis and pleiotropy testing.
Because \citet{keller2018genetic} reported that some nonlocal traits that map to
the Chromosome 2 hotspot did not demonstrate evidence of mediation by \emph{Hnf4a} expression
levels, we elected to study a collection of local gene expression traits,
rather than \emph{Hnf4a} alone.
This strategy enabled us to ask whether one of twelve other local traits mediates those
nonlocal hotspot traits that are not mediated by \emph{Hnf4a}.
Our set of local gene expression traits includes \emph{Hnf4a} and 12
other local genes (Table \ref{tab:annot}).
Our 13 local genes are the only genes that met three criteria (Frame~\ref{frame2}).
The 147 nonlocal traits that we studied all had LOD peak heights above 7.18 and QTL positions within 2 Mb of the center of the hotspot (at 165.5 Mb).

\begin{frameenv}{Local gene inclusion criteria}\label{frame2}
\begin{enumerate}
    \item QTL peak with LOD $>$ 40
    \item QTL peak position within 2 Mb of hotspot center (165.5 Mb)
    \item Gene midpoint within 2 Mb of hotspot center (165.5 Mb)
\end{enumerate}
\end{frameenv}

@@

%\subsection{Scatter plot for all 1911 pairs}
Figure \ref{fig:lod-diff-prop-v-lrt-all} is a scatterplot of the results from all pleiotropy tests and mediation analyses.
Each point contains one local expression trait and one nonlocal expression trait.
I see that points with high values of LOD difference proportion tend to have small
values of pleiotropy test statistic, and those points with high values of the
pleiotropy test statistic tend to have small values of LOD difference proportion.

Some points demonstrate large values of pleiotropy test statistic (\emph{eg.}, $>$ 10),
yet still have sizeable LOD difference proportion statistics (Figure~\ref{fig:lod-diff-prop-v-lrt-all}).
One possible explanation for such points is that the univariate LOD (LOD$_1$) value is small,
so that even a modest LOD difference gives rise to a sizeable LOD difference proportion value.
A second possible explanation is that the phenotype variances for both phenotypes in the pair
are large enough to skew the null distribution of the test statistic towards larger values.
To distinguish between these two, I would
examine the LOD difference statistics and obtain bootstrap $p$-values.

In Figure~\ref{fig:lod-diff-prop-v-lrt-all}, I colored red the points that involve \emph{Hnf4a};
points with other local genes are blue.
The most striking feature of the coloring is that many red points have small
values of the pleiotropy test statistics and very high values of LOD difference
proportion. This is what I expect when \emph{Hnf4a} mediates the relationship
between a nonlocal trait and a QTL; namely, \emph{Hnf4a} and a nonlocal trait
share a pleiotropic QTL and conditioning on \emph{Hnf4a} diminishes the LOD
score for the association between nonlocal trait and QTL.

\begin{figure}
    \centering
    \includegraphics[width = 0.7\textwidth]{../keller2018-chr2-hotspot-chtc/Rmd/lod-diff-prop-v-lrt.eps}
    \caption[LOD difference proportion vs. pleiotropy test statistic]{LOD difference proportion against pleiotropy vs. separate QTL test statistics for 1911 pairs of traits.
    Each point represents a single pair. Pairs that involve local expression trait \emph{Hnf4a} are colored red, while all others are colored blue.
    Note the high concentration of red points in the upper left quadrant of the figure.
    These points, with low values of the pleiotropy vs. separate QTL test statistic and
    high values of LOD difference proportion, are consistent with \emph{Hnf4a} transcript
    levels mediating the effect of \emph{Hnf4a} genetic variants affecting nonlocal
    transcript abundances.}
    \label{fig:lod-diff-prop-v-lrt-all}
\end{figure}

@@

To more thoroughly examine the pleiotropy test and mediation analysis
relationships across the 13 local genes,
I created 13 plots of LOD difference proportion against pleiotropy test
statistic (Figures~\ref{fig:hnf4a} and~\ref{fig:nothnf4a-12}).
They reveal common patterns.
First, I see no points in the upper right quadrant of each plot (Figure~\ref{fig:nothnf4a-12}).
This tells us that those nonlocal genes with high values of pleiotropy test statistic
(when paired with the specified local gene) have low values of LOD difference proportion.
Similarly, those nonlocal genes with high values of LOD difference proportion tend
to have small values of the pleiotropy test statistic.
Finally, some trait pairs demonstrate low values of both
the LOD difference proportion and pleiotropy test statistic.
This observation suggests that, for a given local expression trait, the nonlocal trait
is not mediated by the local expression trait yet it shares a pleiotropic locus.
Low power to resolve multiple nearby QTL may give rise to such findings.

In comparing the \emph{Hnf4a} plot (Figure \ref{fig:hnf4a}) with the other
12 plots (Figure \ref{fig:nothnf4a-12}),
I see that none of the 12 plots in Figure \ref{fig:nothnf4a-12} closely resembles Figure \ref{fig:hnf4a}.
\emph{Serinc3}, \emph{Stk4}, \emph{Neurl2}, and \emph{Cdh22} are closest in appearance to the plot of \emph{Hnf4a}.
However, each of \emph{Serinc3}, \emph{Stk4}, \emph{Neurl2}, and \emph{Cdh22}
has very few points with LOD difference proportion above 0.5,
while \emph{Hnf4a} has many points with LOD difference proportion above 0.5.

\begin{figure}
    \centering
    \includegraphics[width = \textwidth]{../keller2018-chr2-hotspot-chtc/Rmd/Hnf4a-lod-diff-prop-v-lrt.eps}
    \caption[LOD difference proportion vs. pleiotropy test statistic.]{Scatter plot of LOD difference proportion against pleiotropy vs.
    separate QTL test statistic for 147 pairs of traits.
    Each pair includes \emph{Hnf4a} and one of the nonlocal gene expression traits that map to the Chromosome 2 hotspot.}
    \label{fig:hnf4a}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width = 0.8\textwidth]{../keller2018-chr2-hotspot-chtc/Rmd/12local-facet_grid-no-strip.eps}
    \caption[LOD difference proportion vs. pleiotropy test statistic from the per-local gene perspective.]{Scatter plots of LOD difference proportion against pleiotropy vs. separate QTL
    test statistic with 147 pairs of traits per panel.
    Each panel includes a local gene expression trait
    (per the label in the upper right quadrant) and one of the 147 nonlocal gene
    expression traits that map to the Chromosome 2 hotspot.}
    \label{fig:nothnf4a-12}
\end{figure}

@@

@@

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

@@

>>>>

\documentclass[oneside]{book}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\linenumbers

\usepackage{bbm}
\usepackage{setspace}
\usepackage[left = 1in, right = 1in, top = 1.5in, bottom = 1in]{geometry}
\usepackage{multirow}
\usepackage{float}
\usepackage{newfloat}
\usepackage{multicol}
%\usepackage[lite]{mtpro2} %contains \widetilde
\usepackage[titletoc]{appendix}
\usepackage{amsmath}
\usepackage{hyperref}
\hypersetup{unicode=true,
%            pdftitle={Chapter 1},
%            pdfauthor={Frederick Boehm},
%            pdfborder={0 0 0},
            colorlinks=true,
            breaklinks=true}
\usepackage[yyyymmdd,hhmmss]{datetime}

\usepackage[
    backend=biber,
    style=chicago-authordate,
%    citestyle=authoryear,
    natbib=true, %citet and citep
    url=true,
    doi=true,
    eprint=true,
    maxbibnames=10,
    backref=true
]{biblatex}
\DeclareLanguageMapping{english}{english-apa}
\addbibresource{research.bib}
\usepackage{graphicx}
\usepackage{color, colortbl}
\definecolor{LightCyan}{rgb}{0.88,1,1}
\usepackage{subcaption}
\usepackage{array}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{longtable}
\usepackage[outdir=./figs/]{epstopdf}
\usepackage{etoolbox} % for messing with page style & title pages
\usepackage{fancyhdr}
\setlength{\headheight}{15.2pt}
\pagestyle{fancyplain}
\fancyhf{}
\lhead{}
\chead{}
\rhead{ \fancyplain{}{\thepage} } % put page number in upper right corner
\rfoot{Compiled on \today\ at \currenttime}
\renewcommand{\headrulewidth}{0pt} % remove top line
\renewcommand{\footrulewidth}{0pt} % remove bottom line
\renewcommand{\headrule}{}

\title{Expression QTL hotspot dissection}
\author{Frederick J. Boehm, Brian S. Yandell, and Karl W. Broman}
\date{\today}

\begin{document}

\maketitle

<<include=FALSE>>=
library(knitr)
opts_chunk$set(
concordance=TRUE,
echo = TRUE,
fig.lp = "../figures/fig-",
warning = FALSE,
message = FALSE)
@

\doublespacing
\tableofcontents
\listoffigures
\listoftables
%\pagestyle{headings} % page numbers appear at bottom of page, centered!
%\markright{\hfill}
\chapter{Introduction}

Genetics studies in model organisms like mice can identify
genomic regions that affect complex traits, such as systolic blood pressure and body weight \citep{sax1923association,soller1976power,lander1989mapping,broman2009guide,jansen2007quantitative}. These genomic regions are called ``quantitative trait loci'' or ``QTL''.
A genome-wide QTL ``scan'' reveals
associations between genotypes and phenotypes by considering each position,
one at a time, as a candidate QTL for the trait of interest.
A region with strong evidence of association with a complex trait, then, defines a QTL (for that trait).
Because nearby markers have correlated
genotypes, a QTL in a two-parent cross often spans multiple megabases in length
and may contain more than a hundred genes.
Identification of the causal gene (for a given complex trait) from among those genes near the QTL is
challenging and may require costly and time-consuming experiments.
The growing need for greater QTL mapping resolution fueled development over the
last two decades of model organism multiparental populations for high-resolution QTL mapping \citep{de2017back,churchill2004collaborative,svenson2012high,huang2012multiparent,shivakumar2018soybean,huang2011analysis,kover2009multiparent,tisne2017identification,stanley2017genetic}.
With experimentalists now measuring tens of thousands of biomolecular traits
in multiparental populations, the systems genetics community needs multivariate
statistical tools to fully examine the large volumes of data \citep{keller2018genetic,chick2016defining}.
Identifying loci that affect multiple traits can aid in identifying biomolecular interactions and in clarifying complex trait genetic architecture. A QTL that affects more than one trait is called a ``pleiotropic'' QTL.
A test of pleiotropy vs. separate QTL is one multivariate statistical tool that will inform
complex trait genetics by enabling researchers to identify the number of
unique QTL in a genomic region of interest.

<<load-broman>>=
library(broman)
@

Due to the breeding scheme, each Diversity Outbred mouse is a highly
heterozygous and essentially unique mosaic of
founders' DNA. Figure presents color-coded chromosomes for a collection of
subjects from a multiparental population like the Diversity Outbred mice. DNA from each founder
line is color-coded. Each pair of vertical bars corresponds to one subject. Each row corresponds to a single generation, with generation labels on the left. Generation number increases from top to bottom.
With each new generation, the Diversity Outbred mouse genomes are shuffled
in smaller and smaller chunks.

@@

One then fits a linear mixed effects model and calculates the log$_{10}$
likelihood at each position. As in two-parent crosses, one summarizes evidence for a QTL by plotting LOD scores across
the genome. Because multiple hypothesis tests are performed, one typically
wants to control family-wise error rate. One strategy to do this involves a
permutation test, much like that in two-parent crosses \citep{churchill1994empirical}.

@@

We continue by stating what it means for one trait to mediate a relationship between a
DNA variant and another trait. To clarify our discussion, we refer to an example from
\citet{chick2016defining} (Figure). \citet{chick2016defining}, in
studying livers of 192 Diversity Outbred mice, found evidence that \emph{Dhtkd1}
transcript levels associated with a Chromosome 2 marker near the \emph{Dhtkd1} gene.
They also found that
the same marker affected DHTKD1 protein concentrations. (Note that DHTKD1 protein is the
product of translation of \emph{Dhtkd1} transcripts.)
As anticipated, mediation analysis, in which the DHTKD1 protein concentrations are
regressed on founder allele dosages (at the Chromosome 2 marker) demonstrated that
\emph{Dhtkd1} transcript levels act as a mediator between DHTKD1 protein concentrations
and founder allele dosages. In fact, the extent of the reduction in association
strength indicates that the primary pathway by which the genetic marker affects DHTKD1
protein concentrations is through \emph{Dhtkd1} transcript levels.

@@

  \begin{enumerate}
\item No unmeasured confounding of the DNA variant-nonlocal transcript levels relationship
\item No unmeasured confounding of the local transcript levels-nonlocal transcript levels relationship
\item No unmeasured confounding of the DNA variant-local transcript levels relationship
\item No local transcript levels-nonlocal transcript levels confounder that is affected by the DNA variant
\end{enumerate}

Our first statistical challenge is due to confounding. To claim that the LOD
difference statistic reflects a causal relationship, four assumptions about
confounding are needed \citep{vanderweele2015explanation},
yet
it is often difficult or impossible to recognize unmeasured confounders. Unmeasured
confounders, as I use the term here, are unmeasured random variables that confound a causal
relationship between two other random variables.
For example, Figure shows a directed graph. Arrows indicate causal
directions. For example, ``QTL'' causes ``M''. ``M'' designates the mediator (local transcript). ``Y'' denotes the nonlocal transcript
level in this setting, while ``C1'', ``C2'', and ``C3'' are covariates that confound,
respectively, the DNA variant - nonlocal transcript level relationship, the local transcript
level - nonlocal transcript level relationship, and the DNA variant - local transcript level
relationship. Failure to measure C1, C2, or C3 leads to violations of Assumption 1,
Assumption 2, or Assumption 3, respectively.

In studies of Diversity Outbred mice, relatedness is a possible confounder,
yet the linear models (Equations~\ref{model1},~\ref{model2},~\ref{model3},~\ref{model4}) fail to account for the complex relatedness patterns
among Diversity Outbred mice.
In fact, the only covariates in our models are for wave membership and sex.
The 400 mice were shipped from the Jackson Laboratory in four waves of 100 mice.
Because each wave was processed at a different time, I include wave membership
as a covariate with the goal of accounting for wave-related effects.
Other sources of confounding, such as batch effects in the phenotyping, may be
unmeasured.

@@

We designed a study to examine the possible roles for mediation analysis and pleiotropy testing.
Because \citet{keller2018genetic} reported that some nonlocal traits that map to
the Chromosome 2 hotspot did not demonstrate evidence of mediation by \emph{Hnf4a} expression
levels, we elected to study a collection of local gene expression traits,
rather than \emph{Hnf4a} alone.
This strategy enabled us to ask whether one of twelve other local traits mediates those
nonlocal hotspot traits that are not mediated by \emph{Hnf4a}.
Our set of local gene expression traits includes \emph{Hnf4a} and 12
other local genes (Table \ref{tab:annot}).
Our 13 local genes are the only genes that met three criteria.
The 147 nonlocal traits that we studied all had LOD peak heights above 7.18 and QTL positions within 2 Mb of the center of the hotspot (at 165.5 Mb).

\begin{enumerate}
    \item QTL peak with LOD $>$ 7.18
    \item QTL peak position within 2 Mb of hotspot center (165.5 Mb)
    \item Gene midpoint within 2 Mb of hotspot center (165.5 Mb)
\end{enumerate}

@@

%\subsection{Scatter plot for all 1911 pairs}
Figure is a scatterplot of the results from all pleiotropy tests and mediation analyses.
Each point contains one local expression trait and one nonlocal expression trait.
I see that points with high values of LOD difference proportion tend to have small
values of pleiotropy test statistic, and those points with high values of the
pleiotropy test statistic tend to have small values of LOD difference proportion.

Some points demonstrate large values of pleiotropy test statistic (\emph{eg.}, $>$ 10),
yet still have sizeable LOD difference proportion statistics (Figure).
One possible explanation for such points is that the univariate LOD (LOD$_1$) value is small,
so that even a modest LOD difference gives rise to a sizeable LOD difference proportion value.
A second possible explanation is that the phenotype variances for both phenotypes in the pair
are large enough to skew the null distribution of the test statistic towards larger values.
To distinguish between these two, I would
examine the LOD difference statistics and obtain bootstrap $p$-values.

In Figure, I colored red the points that involve \emph{Hnf4a};
points with other local genes are blue.
The most striking feature of the coloring is that many red points have small
values of the pleiotropy test statistics and very high values of LOD difference
proportion. This is what I expect when \emph{Hnf4a} mediates the relationship
between a nonlocal trait and a QTL; namely, \emph{Hnf4a} and a nonlocal trait
share a pleiotropic QTL and conditioning on \emph{Hnf4a} diminishes the LOD
score for the association between nonlocal trait and QTL.

@@

To more thoroughly examine the pleiotropy test and mediation analysis
relationships across the 13 local genes,
I created 13 plots of LOD difference proportion against pleiotropy test
statistic (Figures).
They reveal common patterns.
First, I see no points in the upper right quadrant of each plot (Figure).
This tells us that those nonlocal genes with high values of pleiotropy test statistic
(when paired with the specified local gene) have low values of LOD difference proportion.
Similarly, those nonlocal genes with high values of LOD difference proportion tend
to have small values of the pleiotropy test statistic.
Finally, some trait pairs demonstrate low values of both
the LOD difference proportion and pleiotropy test statistic.
This observation suggests that, for a given local expression trait, the nonlocal trait
is not mediated by the local expression trait yet it shares a pleiotropic locus.
Low power to resolve multiple nearby QTL may give rise to such findings.

In comparing the \emph{Hnf4a} plot (Figure ) with the other
12 plots (Figure ),
I see that none of the 12 plots in Figure  closely resembles Figure .
\emph{Serinc3}, \emph{Stk4}, \emph{Neurl2}, and \emph{Cdh22} are closest in appearance to the plot of \emph{Hnf4a}.
However, each of \emph{Serinc3}, \emph{Stk4}, \emph{Neurl2}, and \emph{Cdh22}
has very few points with LOD difference proportion above 0.5,
while \emph{Hnf4a} has many points with LOD difference proportion above 0.5.

@@

<<r install_here>>=
install.packages("here", repos = "https://cloud.r-project.org")
@

@@

This report was generated on \Sexpr{Sys.time()} using the following computational environment and dependencies:

<<r colophon, cache = FALSE>>=
# which R packages and versions?
devtools::session_info()
@

@@

<<r git2r, cache = FALSE>>=
# what commit is this file at?
git2r::repository(here::here())
@
